%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}


\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{listings}

\lstset{language=Python,
        basicstyle=\scriptsize\ttfamily,
        keywordstyle=\bf, %\color{blue}, % I couldn't find a way to make chars both bold and tt
        frame=none,
        stringstyle=\color{blue},
        fancyvrb=true,
        xleftmargin=10pt,xrightmargin=10pt,
        showstringspaces=false}


\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\davide[1]{\nb{DAV}{#1}}
\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand{\commentout}[1]{}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

\begin{document}


%% \title{Automatic generation of JIT compilers for dynamic languages in
%%   .NET\thanks{This work has been partially supported by MIUR EOS DUE -
%%     Extensible Object Systems for Dynamic and Unpredictable Environments and
%%     by the EU-funded project: IST 004779 PyPy (PyPy: Implementing Python in
%%     Python).}}

\title{Faster than C\#: efficient implementation of dynamic languages on
  .NET\thanks{This work has been partially supported by MIUR EOS DUE -
    Extensible Object Systems for Dynamic and Unpredictable Environments and
    by the EU-funded project: IST 004779 PyPy (PyPy: Implementing Python in
    Python).}}

%% Alternative title: Faster than C\#: the future of dynamic languages on .NET


\numberofauthors{3}
\author{
\alignauthor Antonio Cuni\\
       \affaddr{DISI, University of Genova}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Davide Ancona\\
       \affaddr{DISI, University of Genova}\\
       \affaddr{Italy}\\
       \email{davide@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\conferenceinfo{ICOOOLPS}{'09 Genova, Italy}
\CopyrightYear{2009}
\crdata{978-1-60558-541-3/09/07}

\maketitle

\category{D.3.4}{Programming Languages}{Processors}[code generation,
incremental compilers, optimization, interpreters, run-time environments]

\begin{abstract}
The Common Language Infrastructure (CLI) is a virtual machine expressly
designed for implementing statically typed languages such as C\#, therefore
programs written in dynamically typed languages are typically much slower than C\# when executed on .NET.

Recent developments show that \emph{Just In Time} (JIT) compilers can exploit runtime type
information to generate quite efficient code.  Unfortunately, writing a JIT
compiler is far from being simple.  

In this paper we report our positive
experience with automatic generation of JIT compilers as supported by the PyPy
infrastructure, by focusing on JIT compilation for .NET.
Following this approach, we have in fact added a second layer of JIT compilation, by allowing dynamic generation of more efficient .NET bytecode, which
in turn can be compiled to machine code by the .NET JIT compiler.   

The main and novel contribution of this paper is to show that this
\emph{two-layers JIT} technique is effective, since programs written in dynamic languages 
can run on .NET as fast as (and in some cases even faster than) the equivalent C\# programs.

The practicality of the approach is demonstrated by showing some promising
experiments done with benchmarks written in a simple dynamic language.
\end{abstract}

\input{intro}
\input{clibackend}
\input{benchmarks}
\input{conclusion}




\bigskip

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

% LocalWords:  JIT PyPy
