
\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}

\newcommand{\mynote}[2]{%
  \textcolor{red}{%
    \fbox{\bfseries\sffamily\scriptsize#1}%
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
  }%
}

\newcommand\cfbolz[1]{\mynote{Carl Friedrich}{#1}}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish,
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers,
                                  % short abstracts)

%% \titlebanner{banner above paper title}        % These are ignored unless
%% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{A Way Forward in Parallelising Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

\authorinfo{Remigius Meier}
           {Department of Computer Science\\ ETH Zürich}
           {remi.meier@inf.ethz.ch}
\authorinfo{Armin Rigo}
           {www.pypy.org}
           {arigo@tunes.org}

\maketitle

\begin{abstract}
....
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
%% \terms
%% term1, term2

\keywords
...

\section{Introduction}
...


\section{Conclusion}


%% \appendix
%% \section{Appendix Title}

%% This is the text of the appendix, if you need one.

\acks
...

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem{dan07}
  Dan Grossman. 2007. The transactional memory / garbage collection
  analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
    conference on Object-oriented programming systems and
    applications} (OOPSLA '07).

\bibitem{webjython}
  The Jython Project, \url{www.jython.org}

\bibitem{odaira14}
  Odaira, Rei, Jose G. Castanos, and Hisanobu Tomari.  "Eliminating
  global interpreter locks in Ruby through hardware transactional
  memory."  \emph{Proceedings of the 19th ACM SIGPLAN symposium on
    Principles and practice of parallel programming.} ACM, 2014.

\bibitem{warmhoff13}
  Wamhoff, Jons-Tobias, et al. "FastLane: improving performance of
  software transactional memory for low thread counts."
  \emph{Proceedings of the 18th ACM SIGPLAN symposium on Principles
    and practice of parallel programming.} ACM, 2013.

\bibitem{drago11}
  Dragojević, Aleksandar, et al. "Why STM can be more than a research
  toy." \emph{Communications of the ACM} 54.4 (2011): 70-77.

\bibitem{cascaval08}
  Cascaval, Calin, et al. "Software transactional memory: Why is it
  only a research toy?." \emph{Queue} 6.5 (2008): 40.

\bibitem{nicholas06}
  Nicholas Riley and Craig Zilles. 2006. Hardware transactional memory
  support for lightweight dynamic language evolution. \emph{In
    Companion to the 21st ACM SIGPLAN symposium on Object-oriented
    programming systems, languages, and applications} (OOPSLA
  '06). ACM, New York, NY, USA

\bibitem{fuad10}
  Fuad Tabba. 2010. Adding concurrency in python using a commercial
  processor's hardware transactional memory support. \emph{SIGARCH
  Comput. Archit. News 38}, 5 (April 2010)

\bibitem{felber07}
  Felber, Pascal, et al. "Transactifying applications using an open
  compiler framework." \emph{TRANSACT}, August (2007): 4-6.

\bibitem{bill06}
  Bill McCloskey, Feng Zhou, David Gay, and Eric
  Brewer. 2006. Autolocker: synchronization inference for atomic
  sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
  symposium on Principles of programming languages (POPL '06)}. ACM,
  New York, NY, USA

\bibitem{spear09}
  Spear, Michael F., et al. "Transactional mutex locks." \emph{SIGPLAN
    Workshop on Transactional Computing.} 2009.

\bibitem{lamport79}
  Lamport, Leslie. "How to make a multiprocessor computer that
  correctly executes multiprocess programs." \emph{Computers, IEEE
    Transactions} on 100.9 (1979): 690-691.

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
  transactional memory vs. locks in practice. In \emph{Proceedings of
    the twenty-third annual ACM symposium on Parallelism in algorithms
    and architectures} (SPAA '11). ACM, New York, NY, USA

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. Is transactional programming actually
  easier?. \emph{SIGPLAN} Not. 45, 5 (January 2010), 47-56.

\bibitem{tim03}
  Tim Harris and Keir Fraser. 2003. Language support for lightweight
  transactions. \emph{In Proceedings of the 18th annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '03).

\bibitem{tim05}
  Tim Harris, Simon Marlow, Simon Peyton-Jones, and Maurice
  Herlihy. 2005. Composable memory transactions. \emph{In Proceedings
    of the tenth ACM SIGPLAN symposium on Principles and practice of
    parallel programming} (PPoPP '05).

\bibitem{shan08}
  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. Learning
  from mistakes: a comprehensive study on real world concurrency bug
  characteristics. \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
  329-339.

\bibitem{leis14}
  Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
  Hardware Transactional Memory in Main-Memory Databases."
  \emph{Proc. of ICDE}. 2014.

\bibitem{biased}
  Kenneth Russell and David Detlefs. 2006. Eliminating
  synchronization-related atomic operations with biased locking and
  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '06).

\end{thebibliography}


\end{document}
